---
title: "Statistical Analysis and visualization of Kalmia fruits"
author: "Callin Switzer"
date: "February 10, 2017"
output: html_document
---


<pre>
# Kalmia analyze fruit size at end of season
# Using fruit sizes calculated from image segmentation in Python with opencv
#  Callin Switzer
# 20 October 2016
# 10 Feb 2017 Update: Conducted Statistical Modeling with LMER and GLMER
# 11 March 2017 Update: added random effect to MER models to account for plant lineage
</pre>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/callinswitzer/Dropbox/ExperSummer2016/Kalmia/Manuscript/DatasetsSupplemental/")
```


## Read in data

```{r, warning=FALSE, message = FALSE}
ipak <- function(pkg){
     new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
     if(length(new.pkg)) install.packages(new.pkg, dependencies = TRUE)
     sapply(pkg, require, character.only = TRUE)
}

packages <- c("ggplot2", 'lme4', 'plyr', 'influence.ME', 'sjPlot', 'multcomp', 'lsmeans', 'Matrix')
ipak(packages)

setwd("/Users/callinswitzer/Dropbox/ExperSummer2016/Kalmia/Manuscript/DatasetsSupplemental/")

theme_set(theme_classic())

kfrt <- read.csv("kalmiaFruitFinal.csv", stringsAsFactors = FALSE)
nrow(kfrt[kfrt$dia_mm == 20.0,]) # number of images taken

# clean and process data
kfrt <- kfrt[kfrt$dia_mm != 20.0, ] # 20 is the reference object in segmented images
nrow(kfrt) # number of total fruits measured


# label treatmens and access numbers
kfrt$trt <- sapply(X = 1:nrow(kfrt), FUN = function(x) strsplit(kfrt$plantNum[x], split = "__")[[1]][2])
kfrt$accessNum <- sapply(X = 1:nrow(kfrt), FUN = function(x) strsplit(kfrt$plantNum[x], split = "__")[[1]][1])

kfrt$trt <- mapvalues(kfrt$trt, c(1,2,3,4,5), c("Bagged", "Bagged & Selfed", "Unbagged", 
                                                "Unbagged & Outcrossed", "Unbagged_2"))
# add plant lineage (all plants that start with 1129 are from the lineage, 673_69 )
plantInds <- 1:nrow(kfrt) %in% grep(pattern = "1129", x = kfrt$plantNum)
kfrt$plantLineage <- sapply(1:length(plantInds), FUN = function(x){
     ifelse(plantInds[x], "673_69", paste(strsplit(kfrt$plantNum[x], split = "_")[[1]][1:2], collapse = "_") )})
```


# Add plants with 0 count to the dataset
```{r}
# count number of fruits
counts = as.data.frame(table(kfrt$plantNum))
counts$trt = sapply(X = 1:nrow(counts), FUN = function(x) strsplit(as.character(counts$Var1[x]), split = "__")[[1]][2])

counts$trt <- mapvalues(counts$trt, c(1,2,3,4,5), c("Bagged", "Bagged & Selfed", "Unbagged", 
                                                "Unbagged & Outcrossed", "Unbagged_2"))


counts$accessNum = sapply(X = 1:nrow(counts), FUN = function(x) strsplit(as.character(counts$Var1[x]), split = "__")[[1]][1])


# add in the trts and accession numbers that had counts of 0
# read in error-checked datasheet
kalNotes <- read.csv("KalmiaDailyDatasheets_ErrorChecked.csv")

# get the accession numbers I should have
accNumsHave <- unique(kalNotes$Plant.Number)
accNumsHave <- accNumsHave[!(accNumsHave %in% c("", "Plant Number"))]

#change formatting to match above
accNumsHave <- gsub("#", "", accNumsHave)
accNumsHave <- gsub("\ ", "_", accNumsHave)
accNumsHave <- gsub("\\-", "_", accNumsHave)
accNumsHave <- gsub("\\*", "_", accNumsHave)
accNumsHave <-toupper(accNumsHave)

shouldHave <- as.data.frame(t(sapply(accNumsHave, function(x) as.character(paste(x, c(1:4), sep = "__")))))
row.names(shouldHave) <- NULL
shouldHave1 <- c(as.character(shouldHave[, 1]), 
                 as.character(shouldHave[, 2]), 
                 as.character(shouldHave[, 3]), 
                 as.character(shouldHave[, 4]))

# find missing ones -- this is the ones that
# are in the daily datasheet that aren't in the fruit measurements
missingTrts <- setdiff(shouldHave1,unique(kfrt$plantNum[kfrt$trt != "5"]))

# this should be 0 -- the ones from the fruit measurements that aren't in the daily datasheet
setdiff(unique(kfrt$plantNum[kfrt$trt != "Unbagged_2"]), shouldHave1)

# here's the list of plants that had their bags/tags go missing during the experiment (meaning that
# I was unable to collect fruits)
# note: "677_66_MASS__1" was the only plant that was missing a tag during the final collection
# that wasn't missing in one of my previous checks. 
NAList <- c("1129_74_E__4", "1129_74_C__4", "39_60_A__3", "685_70_A__2", "677_66_MASS__1")

# note: this ignores trt#5 which is the same as #3
ZeroFruitPlants <- missingTrts[!(missingTrts %in% NAList)]

zfdf <- data.frame(Var1 = ZeroFruitPlants, 
                   Freq = 0, 
                   trt = sapply(X = 1:length(ZeroFruitPlants), 
                                FUN = function(x) strsplit(as.character(ZeroFruitPlants[x]), 
                                                           split = "__")[[1]][2]),
                   accessNum = sapply(X = 1:length(ZeroFruitPlants), 
                                            FUN = function(x) strsplit(as.character(ZeroFruitPlants[x]), 
                                                                       split = "__")[[1]][1])
                   )

zfdf$trt <- mapvalues(zfdf$trt, c(1,2,3,4,5), c("Bagged", "Bagged & Selfed", "Unbagged", 
                                                "Unbagged & Outcrossed", "Unbagged_2"))

countFin <- rbind(counts, zfdf)


# final fruit counts for the fruits collected at the end of the experiment
countFin <- countFin[order(countFin$accessNum, countFin$trt, decreasing = FALSE), ]

# change labels from unbagged to conntrol
countFin$trt <- mapvalues(countFin$trt, c("Bagged", "Bagged & Selfed", "Unbagged", 
                                                "Unbagged & Outcrossed", "Unbagged_2")
                          , c("Bagged", "Bagged & Selfed", "Control", 
                                                "Unbagged & Outcrossed", "Control_2"))
# change levels, so that control is first
countFin$trt <- factor(countFin$trt, levels = c("Control","Control_2", "Bagged", "Bagged & Selfed", 
                                                "Unbagged & Outcrossed"))

plantInds <- 1:nrow(countFin) %in% grep(pattern = "1129", x = countFin$accessNum)
countFin$plantLineage <- sapply(1:length(plantInds), FUN = function(x){
     ifelse(plantInds[x], "673_69", paste(strsplit(countFin$accessNum[x], split = "_")[[1]][1:2], collapse = "_") )})

```

# Visualize counts of fruits
```{r}
ggplot(countFin, aes(x = trt , y = Freq, fill = trt)) + 
     geom_boxplot() +
    # geom_violin()+
     
     labs(y = "Number of fruits", x = "Treatment") + 
     scale_fill_brewer(name = "Treatment", palette = "Set1")
saveDir <- "/Users/callinswitzer/Dropbox/ExperSummer2016/Kalmia/Manuscript/Media/"
ggsave(paste0(saveDir, "KalmiaFruitNumber.pdf"), width = 10, height = 8)

```


# Visualize counts of fruits (ignore treatment #5)
```{r}
ggplot(countFin[!(countFin$trt %in% 'Control_2'), ], aes(x = trt , y = Freq, fill = trt)) + 
     geom_boxplot() +
     # geom_point(aes(color = plantLineage)) + 
     labs(y = "Number of fruits", x = "Treatment") +
     theme(axis.text.x = element_text(angle = 35, hjust = 1), 
           legend.position = 'none') + 
     scale_fill_brewer(name = "Treatment", palette = "Set1")

ggsave(paste0(saveDir, "KalmiaFruitNumber_trt1_4Only.pdf"), width = 5, height = 4)
```

# Use GLMM's to find differences in number of fruits
### First, summarize the dataset that I will be using
```{r}
# I'm ignoring control#2, because it was originally intended for another experiment (analysis wasn't planned)
cf1 <- droplevels(countFin[countFin$trt != "Control_2",])
nrow(countFin) # number of total counts, including Control2
sum(cf1$Freq) # total number of fruits in the analysis of count
# Number of counts, excluding control_2
# this is different from the number of photos
# because I didn't take photos on 0-count plants
nrow(cf1) 
data.frame(table(cf1$trt)) # number of plants in each treatment that we are analyzing
data.frame(table(cf1$accessNum)) # shows number of counts per plant -- 4 means that we counted all four treatments
data.frame(table(cf1$plantLineage))


# shows which treatments / plants are missing from analysis
# this is different from the plants that just had 0 counts
data.frame(table(interaction(cf1$accessNum , cf1$trt))) 
```

# Create a model
Account for plant lineage and accession number

```{r}
m1 <- glmer(Freq ~ trt + (1|plantLineage) +  (1|accessNum), family = poisson, data = cf1)
summary(m1)

# calculate LRT for trt
m2 <- update(m1, .~. - trt)
anova(m1, m2) # highly significant
```


# Model diagnostics
```{r, cache = TRUE}
m1 <- glmer(Freq ~ trt + (1|plantLineage) +  (1|accessNum), family = poisson, data = cf1)
summary(m1)

# diagnostics
# qq plot
qqnorm(resid(m1), main = "")
qqline(resid(m1)) # outliers

# residual plot
plot(fitted(m1), resid(m1), xlab = "fitted", ylab = "residuals")
abline(0,0)
# possibly outliers


# QQPlot for group-level effects
qqnorm(ranef(m1)$accessNum[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(m1)$accessNum[[1]]) 

# # QQPlot for group-level effects
qqnorm(ranef(m1)$plantLineage[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(m1)$plantLineage[[1]])

infl <- influence(m1, obs = TRUE) # takes a while to calculate
# x11()
plot(infl, which = 'cook') # some influential points

# visualize model: 
sjp.lmer(m1, type = 'fe')
sjp.lmer(m1, type = 're', sort = TRUE) # plot random effects to look for outliers

#check assumptions of model 
overdisp_fun <- function(model) {
  ## number of variance parameters in 
  ##   an n-by-n variance-covariance matrix
  vpars <- function(m) {
    nrow(m)*(nrow(m)+1)/2
  }
  model.df <- sum(sapply(VarCorr(model),vpars))+length(fixef(model))
  rdf <- nrow(model.frame(model))-model.df
  rp <- residuals(model,type="pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq/rdf
  pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
  c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

overdisp_fun(m1) # shows overdispersion

# here's another way to check for overdispersion
residDev <- sum(residuals(m1, type = 'deviance')^2) # calculate residual deviance
# this ratio should be about 1 -- larger than 1 suggests overdispersion
residDev / df.residual(m1) 
```


## Use negative binomial model, since the poisson model is overdispersed
```{r}
m1.1 <- glmer.nb(Freq ~ trt +(1|plantLineage) +  (1|accessNum), data = cf1)
summary(m1.1)


# get overall p-value for treatment
m2.1 <- update(m1.1, .~. - trt)
anova(m1.1, m2.1, test = 'LRT') # LRT for trt for paper
```

# Negative binomial model diagnostics
```{r, cache = TRUE}
overdisp_fun(m1.1) 

# here's another way to check for overdispersion
residDev <- sum(residuals(m1.1, type = 'deviance')^2) # calculate residual deviance
# this ratio should be about 1 -- larger than 1 suggests overdispersion
residDev / df.residual(m1.1) 

# diagnostics
# qq plot
qqnorm(resid(m1.1), main = "")
qqline(resid(m1.1)) # a little better

# residual plot
plot(fitted(m1.1), resid(m1.1), xlab = "fitted", ylab = "residuals")
abline(0,0) # residuals look better for this model


# QQPlot for group-level effects
qqnorm(ranef(m1.1)$accessNum[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(m1.1)$accessNum[[1]])

qqnorm(ranef(m1.1)$plantLineage[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(m1.1)$plantLineage[[1]]) 

infl <- influence(m1.1, obs = TRUE) # takes a while to calculate
plot(infl, which = 'cook') # nothing above 1

# visualize model: 
sjp.lmer(m1.1, type = 'fe')
sjp.lmer(m1.1, type = 're', sort = TRUE) # plot random effects to look for outliers

# get estimated dispersion parameter for NB Model
getME(m1.1, "glmer.nb.theta") # 2.17

# post-hoc pairwise comparisons with adjusted p-values
# from documentation: 
# test = adjusted() multiple test procedures as specified by the type argument
# to adjusted: "single-step" denotes adjusted p values as computed
# from the joint normal or t distribution of the z statistics (default),
summary(glht(m1.1, lsm(pairwise ~ trt)), test=adjusted("single-step")) # pairwise tests for fruit counts
summary(glht(m1.1, lsm(pairwise ~ trt)), test=adjusted("none")) # pairwise tests for fruit counts, unadjusted
```

# Visualize annotated plot for fruit counts
```{r}
count_pub <- within(countFin, {
     trt <- mapvalues(trt, from = c("Control","Control_2", "Bagged", "Bagged & Selfed", 
                                                "Unbagged & Outcrossed"), 
                      to = c("Control","Control_2", "Bagged", "Bagged\n & Selfed", 
                                                "Outcrossed"))
})

count_pub <- droplevels(count_pub)


ggplot(count_pub[!(count_pub$trt %in% 'Control_2'), ], aes(x = trt , y = Freq+ 0.1)) + 
     geom_boxplot(width = 0.2, fill = 'grey80') +
     labs(y = "Number of fruits", x = "Treatment") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           legend.position = 'none') + 
     # annotate
     annotate(geom="text", x=c(1,2,3,4), y=c(70, 70, 70, 70) + 10, label=c("a", "b", "a", "c"),
               color="black") + 
     scale_y_log10(breaks = c(1, 10, 100), limits = c(0.1, 100))

ggsave(paste0(saveDir, "KalmiaFruitNumber_logScale.pdf"), width = 3, height = 4)
```


# Visualize CI's for each of the groups
```{r, cache = TRUE}
# refit m1.1 with treatments in alphabetical order (b/c order isn't preserved in model matrix)
cf1.1 <- within(cf1, {
     trt = as.factor(as.character(cf1$trt))
})

# refit model with different reference levels
m1.2 <- glmer.nb(Freq ~ trt + (1|plantLineage) + (1|accessNum), data = cf1.1)
summary(m1.2)

# may help solve convergence issues-- NOPE!
pframe <- data.frame(trt = levels(droplevels(cf1.1$trt)), plantLineage = 99999)
pframe$Freq <- 0
pp <- predict(m1.2, newdata = pframe, re.form=NA, type = 'response') # re.form sets all random effects to 0


### Calculate CI's (using bootstrap, not accounting for random effects)
## I think that sometimes the model doesn't converge, because it estimates variance
## for random effects to be 0.
##  according to this: http://rstudio-pubs-static.s3.amazonaws.com/24365_2803ab8299934e888a60e7b16113f619.html
## you can safely ignore the warnings about convergence

bb2 <- bootMer(m1.2, FUN=function(x) predict(x, pframe, re.form=NA, type = 'response'), nsim = 999)
bb2_se <-apply(bb2$t,2,function(x) quantile(x, probs = c(0.025, 0.975)))
pframe$blo<-bb2_se[1,]
pframe$bhi<-bb2_se[2,]
pframe$predMean <- pp


# This method is the same, but less straightforward.
# mm <- model.matrix(terms(m1.2), pframe)
# predFun<-function(.) exp(mm%*%fixef(.) )
# bb<-bootMer(m1.2,FUN=predFun,nsim=200) #do this 200 times
# 
# # as we did this 200 times the 95% CI will be bordered by the 5th and 195th value
# bb_se<-apply(bb$t,2,function(x) quantile(x, probs = c(0.025, 0.975)))
# pframe$blo<-bb_se[1,]
# pframe$bhi<-bb_se[2,]
# pframe$predMean <- pp

pframe$median = tapply(cf1.1$Freq, INDEX = cf1.1$trt, median)
pframe$trt <- mapvalues(pframe$trt, from = c("Control", "Bagged", "Bagged & Selfed", 
                                                "Unbagged & Outcrossed"),
                        to = c("Control", "Bagged", "Bagged\n & Selfed", 
                                                "Outcrossed"))
pframe$trt <- relevel(pframe$trt, ref = "Control")

#plot 95% confidence intervals
# "Mean and bootstrap CI based on fixed-effects uncertainty ONLY"
number_ticks <- function(n) {function(limits) pretty(limits, n)}
g0 <- ggplot(pframe, aes(x=trt, y=predMean))+
     geom_point()+ 
     labs(y = "Number of fruits", x = "Treatment") + 
     geom_errorbar(aes(ymin = blo, ymax = bhi), width = 0.1)+
     scale_y_log10(limits =c(1,60), breaks = number_ticks(6)) + 
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           legend.position = 'none') +
     # annotate
     annotate(geom="text", x=c(1,2,3,4), y=c(70, 70, 70, 70) - 10, label=c("a", "b", "a", "c"),
               color="black") 
g0
ggsave(paste0(saveDir, "KalmiaFruitNumber_BSCI_logScale.pdf"), width = 3, height = 4)


## Bootstrap CI on linear scale -- not that great!
g1 <- ggplot(pframe, aes(x=trt, y=predMean))+
     geom_point()+ 
     labs(y = "Number of Fruits", x = "Treatment") + 
     geom_errorbar(aes(ymin = blo, ymax = bhi), width = 0.1)+ 
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           legend.position = 'none') +
     # annotate
     annotate(geom="text", x=c(1,2,3,4), y=c(70, 70, 70, 70) - 40, label=c("a", "b", "a", "c"),
               color="black") 
g1
ggsave(paste0(saveDir, "KalmiaFruitNumber_BSCI_LinearScale.pdf"), width = 3, height = 4)


cp <- droplevels(count_pub[count_pub$trt != 'Control_2',])

# this might actually be the best way
ggplot(cp, aes(x = trt , y = Freq)) + 
     geom_boxplot(width = 0.2, fill = 'grey80') +
     labs(y = "Number of fruits", x = "Treatment") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           legend.position = 'none') + 
     # annotate
     annotate(geom="text", x=c(1,2,3,4), y=c(70, 70, 70, 70) + 1, label=c("a", "b", "a", "c"),
               color="black") 
ggsave(paste0(saveDir, "KalmiaFruitNumber_LinearScale.pdf"), width = 3, height = 4)
```


# Visualize fruit size
```{r}
# calculate fruit size by plant
sizeDF_mean <- as.data.frame(tapply(kfrt$dia_mm, INDEX = kfrt$plantNum, mean))
colnames(sizeDF_mean) = "meanFrtSz"
sizeDF_mean$trt = sapply(X = 1:length(sizeDF_mean$meanFrtSz), 
                          FUN = function(x) strsplit(as.character(row.names(sizeDF_mean)[x]), 
                                                     split = "__")[[1]][2])

sizeDF_mean$trt <- mapvalues(sizeDF_mean$trt, c(1,2,3,4,5), c("Bagged", "Bagged & Selfed", "Control", 
                                                "Unbagged & Outcrossed", "Control_2"))

# reorder
sizeDF_mean$trt <- factor(sizeDF_mean$trt, levels = c("Control", "Control_2", "Bagged", "Bagged & Selfed",  
                                                "Unbagged & Outcrossed"))

sizeDF_mean$accessNum = sapply(X = 1:length(sizeDF_mean$meanFrtSz), 
                   FUN = function(x) strsplit(as.character(row.names(sizeDF_mean)[x]), 
                                              split = "__")[[1]][1])


ggplot(sizeDF_mean, aes(x = trt, y = meanFrtSz, fill = trt)) + 
     geom_boxplot(alpha = 0.5) + 
#      stat_summary(fun.y=mean, geom="line", aes(group = accessNum, color = accessNum))  + 
#      stat_summary(fun.y=mean, geom="point", aes(group = accessNum, color = accessNum)) + 
     geom_point(aes(color = trt))


ggplot(sizeDF_mean, aes(x = trt, y = meanFrtSz, fill = trt)) + 
     geom_boxplot() + 
     labs(y = "Mean Fruit Diameter (mm)", x = "Treatment") + 
     scale_fill_brewer(name = "Treatment", palette = "Set1")
ggsave(paste0(saveDir, "KalmiaFruitDiameter.pdf"), width = 10, height = 8)


# visualize , but exclude treatment 5
ggplot(sizeDF_mean[!(sizeDF_mean$trt %in% 'Control_2'), ], 
       aes(x = trt, y = meanFrtSz, fill = trt)) + 
     geom_boxplot() + 
     labs(y = "Mean Fruit Diameter (mm)", x = "Treatment") + 
     scale_fill_brewer(name = "Treatment", palette = "Set1") + 
     theme(axis.text.x = element_text(angle = 35, hjust = 1), 
          legend.position = 'none') 
ggsave(paste0(saveDir, "KalmiaFruitDiameter_trt14Only.pdf"), width = 5, height = 4)
```


# Model Fruit Size with LMER
```{r}
size1 <- within(kfrt, {
     trt <- mapvalues(trt, c("Bagged", "Bagged & Selfed", "Unbagged", 
                                                "Unbagged & Outcrossed", "Unbagged_2"), 
                      c("Bagged", "Bagged\n & Selfed", "Control", 
                                                "Outcrossed", "Control_2"))
})

size1 <- droplevels(size1[size1$trt != "Control_2",])

# get sample sizes for size dataset
nrow(size1) # total number of fruits in analysis for size
data.frame(table(size1$accessNum)) # total number of fruits per plant

data.frame(table(size1$plantLineage)) # total number of fruits per plant lineage


size1$trt <- relevel(as.factor(size1$trt), ref = "Control")
f1 <- lmer(dia_mm ~ trt + (1|plantLineage/accessNum), data = size1)
summary(f1) # final model for paper

f1 <- lmer(dia_mm ~ trt + (1|plantLineage) + (1|accessNum), data = size1)
summary(f1) # final model for paper (same as above)

# get p-value for trt
f2 <- update(f1, .~. - trt)
anova(f1, f2, "LRT")
```


# Fruit Size Diagnostics
```{r, cache = TRUE}
# diagnostics
# qq plot
qqnorm(resid(f1), main = "")
qqline(resid(f1)) # good

# residual plot
plot(fitted(f1), residuals(f1, type = "deviance"), xlab = "fitted", ylab = "residuals")
abline(0,0) 


# QQPlot for group-level effects
qqnorm(ranef(f1)$accessNum[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(f1)$accessNum[[1]]) # one possible outlier

# QQPlot for group-level effects
qqnorm(ranef(f1)$plantLineage[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(f1)$plantLineage[[1]]) # looks good


infl <- influence(f1, obs = TRUE) # takes a while to calculate
plot(infl, which = 'cook') # nothing above 1

# visualize model: 
sjp.lmer(f1, type = 'fe')

# Best Linear Unbiased Predictors (BLUPs)
sjp.lmer(f1, type = 're', sort = TRUE) # plot random effects to look for outliers

# post-hoc pairwise comparisons with adjusted p-values
# from documentation: 
# test = adjusted() multiple test procedures as specified by the type argument
# to adjusted: "single-step" denotes adjusted p values as computed
# from the joint normal or t distribution of the z statistics (default),
summary(glht(f1, lsm(pairwise ~ trt)), test=adjusted("single-step")) # pairwise tests for fruit counts
summary(glht(f1, lsm(pairwise ~ trt)), test=adjusted("none")) # pairwise tests for fruit counts, unadjusted
```


# Visualize Fruit Size with annotations
```{r}
# refit m1.1 with treatments in alphabetical order (b/c order isn't preserved in model matrix)
sf1<- within(size1, {
     trt = as.factor(as.character(size1$trt))
})

# refit model with different reference levels
f1 <- lmer(dia_mm ~ trt + (1|plantLineage) + (1|accessNum), data = sf1)
pframe <- data.frame(trt = levels(droplevels(sf1$trt)))
pframe$dia_mm = 0
pp <- predict(f1, newdata = pframe, re.form=NA, type = 'response') # re.form sets all random effects to 0


### Calculate CI's (using bootstrap, not accounting for random effects)
mm <- model.matrix(terms(f1), pframe)
predFun<-function(.) mm%*%fixef(.)
bb<-bootMer(f1,FUN=predFun,nsim=999) #do this 200 times
# get quantiles from bootstrap sample
bb_se<-apply(bb$t,2,function(x) quantile(x, probs = c(0.025, 0.975)))
pframe$blo<-bb_se[1,]
pframe$bhi<-bb_se[2,]
pframe$predMean <- pp
pframe # print frame, in case reporting in a table

pframe$trt <- relevel(pframe$trt, ref = "Control")

#plot 95% confidence intervals
# "Mean and bootstrap CI based on fixed-effects uncertainty ONLY"
g2 <- ggplot(pframe, aes(x=trt, y=predMean))+
     geom_point()+ 
     labs(y = "Fruit dia. (mm)", x = "Treatment") + 
     geom_errorbar(aes(ymin = blo, ymax = bhi), width = 0.1)+
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           legend.position = 'none') +
     # annotate
     annotate(geom="text", x=c(1,2,3,4), y=c(5,5,5,5) + 0.7, label=c("a", "b", "a", "c"),
               color="black") 
g2
ggsave(paste0(saveDir, "KalmiaFruitDia_BSCI.pdf"), width = 3, height = 4)

 ## visualize raw data (mean fruit size per plant)

sdf1 <- droplevels(within(sizeDF_mean[!(sizeDF_mean$trt %in% 'Control_2'), ], {
     trt <- mapvalues(trt, c("Bagged", "Bagged & Selfed", "Control", 
                                                "Unbagged & Outcrossed", "Control_2"), 
                      c("Bagged", "Bagged\n & Selfed", "Control", 
                                                "Outcrossed", "Control_2"))
}))

# visualize fruit size
ggplot(sdf1, aes(x = trt, y = meanFrtSz)) + 
     geom_boxplot(width = 0.2, fill = 'grey80') +
     labs(y = "Fruit dia. (mm)", x = "Treatment") +
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           legend.position = 'none') + 
     # annotate
     annotate(geom="text", x=c(1,2,3,4), y=c(5,5,5,5) + 2, label=c("a", "b", "a", "c"),
               color="black")
ggsave(paste0(saveDir, "KalmiaFruitDia.pdf"), width = 3, height = 4)


# compare two plots
g2 + geom_boxplot(data = sdf1, aes(x = trt, y = meanFrtSz), width = 0.2, alpha = 0)
```


# Visualize data to compare fruit size with number of seeds

```{r}
# read in data
sizeSeed <- read.csv("KalmiaFruitSizeAndSeeds.csv")
plantInds <- 1:nrow(sizeSeed) %in% grep(pattern = "1129", x = sizeSeed$Plant)
sizeSeed$plantLineage <- sapply(1:nrow(sizeSeed), FUN = function(x){
     ifelse(plantInds[x], "673_69", paste(strsplit(as.character(sizeSeed$Plant[x]), split = "_")[[1]][1:2], collapse = "_") )})


ggplot(sizeSeed, aes(x = Dia..mm., y = NumSeeds)) + 
     geom_point() + 
     stat_smooth(method = 'lm', formula = y ~ exp(x), se = F) + 
     labs(x = 'Fruit Diameter (mm)', y = 'Num Seeds in 1 Carpel')

ggsave(paste0(saveDir, "KalmiaFruitSeeds.pdf"), width = 5, height = 4)


# on log scale
ggplot(sizeSeed, aes(x = Dia..mm., y = NumSeeds)) + 
     geom_point() + 
     stat_smooth(method = 'lm', formula = y ~ x, se = F) + 
     scale_y_continuous(trans="log") + 
     labs(y = "log(e) number of seeds")

# GLMER
ss1 <- glmer(NumSeeds ~ Dia..mm. + (1|plantLineage), family = poisson, data = sizeSeed)
summary(ss1)

ss2 <- update(ss1, .~. - Dia..mm.)
anova(ss1, ss2, test = "LRT") # LRT for paper


residDev <- sum(residuals(ss1, type = 'deviance')^2) # calculate residual deviance
# this ratio should be about 1 -- larger than 1 suggests overdispersion
residDev / df.residual(ss1) 


overdisp_fun(ss1) #not overdispersed

## model diagnostics 
plot(ss1) #cook's distance

qqnorm(resid(ss1), main = "")
qqline(resid(ss1)) # one outlier


# # QQPlot for group-level effects
qqnorm(ranef(ss1)$plantLineage[[1]], main="Normal Q-Q plot for random effects")
qqline(ranef(ss1)$plantLineage[[1]])

infl <- influence(ss1, obs = TRUE) # takes a while to calculate
plot(infl, which = 'cook') # some influential points

# get sample size
nrow(sizeSeed) # total number of individual fruits

sum(sizeSeed$NumSeeds) # total number of seeds counted

# total plant lineages
length(unique(sizeSeed$plantLineage))


# plot data with line
preds <- predict(ss1, newdata = data.frame(Dia..mm. = seq(2.5, 5.5, length.out = 100)), re.form = NA, type = 'response')

predDF <- data.frame(Dia..mm. = seq(2.5, 5.5, length.out = 100), preds = preds)


# visualize data
ggplot(sizeSeed, aes(x = Dia..mm., y = NumSeeds)) + 
     geom_point() + 
     geom_line(data = predDF, aes(x = Dia..mm., y = preds), color = 'blue') +  
     labs(x = 'Fruit Diameter (mm)', y = 'Num Seeds in 1 Carpel')

ggsave(paste0(saveDir, "KalmiaFruitDia_NumSeeds.pdf"), width = 3, height = 4)
```


### Print sesion info
```{r}
sessionInfo()

# print time of that html document was created
Sys.time()
```